---
title: "Time Use Exploration"
author: "Mallick Hossain"
output:
  html_document:
    code_folding: hide
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: no
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(cache.path = "../../../../../Desktop/TimeData/cache")
library(data.table)
library(bit64)
library(plotly)
library(purrr)
```

# Introduction 

This report will explore a variety of trends in American time use that have 
occurred over the past 12 years, which has been gathered using the American 
Time Use Survey published by the Bureau of Labor Statistics. In particular, this
report will start off by introducing a wide range of trends that can be seen in 
the data. Later in the report, we will explore various explanations for the 
underlying causes of these trends. 

# Data Description

The American Time Use Survey (ATUS) is a nationally representative survey of how
Americans spend their time. It began in January 2003 and has been collected
regularly since then. Households that complete their final (8th) month of the
Current Population Survey are eligible for the ATUS. Households are selected to
represent a range of demographics and within that household, one person age 15
or older is randomly chosen. That person is interviewed between 2-5 months after
the household's final CPS interview (majority are interviewed within 3 months.
The interview asks respondents to collect a detailed account of their activities
(including exact time intervals) over the 24 hour period from 4am of the previous
day to 4am of the interview day. For each activity, the duration, location, and
other participants are also recorded. Activities are classified across over 400
different detailed categories.

The 2003 wave of the ATUS has information on over 20,000 respondents and each of
the successive waves has included about 13,000 respondents.

# Summary Statistics
In this section, we look at some summary statistics related to how time use varies
by gender, location, and other groups. First I load and merge the relevant data
(namely activity level data with the demographics of each respondent) and create 
indicators for home/non-home locations and age groups.
```{r loadMerge, echo = FALSE, include = FALSE, eval = FALSE}
# Loading the data
timeUse <- fread("~/Desktop/TimeData/atusact_0315/atusact_0315.dat")
demographics <- fread("~/Desktop/TimeData/atussum_0315/atussum_0315.dat")

# Dropping activity columns from demographics data table (already contained in
# timeUse data table)
activityCols <- grep("t", names(demographics))
demographics[, (activityCols) := NULL]

# Merging data
setkey(timeUse, "TUCASEID")
setkey(demographics, "TUCASEID")
mergedData <- timeUse[demographics]
timeUse <- mergedData
rm(mergedData, demographics, activityCols)

# Generating years, home dummy, and age group
timeUse[, `:=`(TUYEAR = as.integer(substr(TUCASEID, 1, 4)),
              home = ifelse(TEWHERE == -1, -1,
                     ifelse(TEWHERE == -2, -2,
                     ifelse(TEWHERE == -3, -3,
                     ifelse(TEWHERE == 1, 1, 2)))),
              ageGroup = ifelse(TEAGE >= 18 & TEAGE <= 24, 1,
                         ifelse(TEAGE >= 25 & TEAGE <= 54, 2,
                         ifelse(TEAGE >= 55 & TEAGE <= 64, 3,
                         ifelse(TEAGE >= 65, 4, NA)))))]

# Labeling gender
timeUse$TESEX <- factor(timeUse$TESEX, levels = c(1, 2),
                        labels = c("Male", "Female"))

# Labeling location
timeUse$home <- factor(timeUse$home, levels = c(-1, -2, -3, 1, 2),
                        labels = c("Blank", "Don't Know", "Refused", "Home", "Away"))

# Labeling age groups
timeUse$ageGroup <- factor(timeUse$ageGroup, levels = c(1, 2, 3, 4),
                           labels = c("18-24", "25-54", "55-64", "65+"))

# Labeling metro areas
timeUse$GTMETSTA <- factor(timeUse$GTMETSTA, levels = c(-1, 1, 2, 3),
                           labels = c("Blank", "Metro", 
                                      "Non-Metro", "Not Identified"))

# Labeling 2 digit activity categories. Note that converting to factors
# changes the levels to from 1 to 18 instead of from 1 to 16, 18, and 50.
timeUse$TRTIER1P <- factor(timeUse$TRTIER1P, levels = c(1:16, 18, 50),
                           labels = activities <- c("Personal Care",
                                                    "Household Activities",
                                                    "Caring for and Helping HH Members",
                                                    "Caring for and Helping non-HH Members",
                                                    "Work and Related Activities",
                                                    "Education",
                                                    "Consumer Purchases",
                                                    "Professional and Personal Care Services",
                                                    "Household Services",
                                                    "Government Services and Civics",
                                                    "Eating and Drinking",
                                                    "Socializing, Relaxing, and Leisure",
                                                    "Sports, Exercise, and Recreation",
                                                    "Religion",
                                                    "Volunteering",
                                                    "Telephone Calls",
                                                    "Traveling",
                                                    "Unknown"))

save(timeUse, file = "~/Desktop/TimeData/data/timeUseMerged.rda", compress = TRUE)
rm(list = ls())
```

Then, we can clean the data to obtain the groups we are concerned with. I remove
any repondents younger than 18
```{r filterData, echo = FALSE, include = FALSE, eval = FALSE}
load(file = "~/Desktop/TimeData/data/timeUseMerged.rda")

# Dropping observations of those younger than 18 (7,541 respondents)
before <- length(unique(timeUse$TUCASEID))
timeUse <- timeUse[TEAGE >= 18]
after <- length(unique(timeUse$TUCASEID))
before - after

## Dropping observations that were missing, can't remember, or other insufficient data (removing 30,501 activity obs)
#timeUse <- timeUse[TRTIER1P != 50]

save(timeUse, file = "~/Desktop/TimeData/data/timeUseClean.rda", compress = TRUE)
rm(list = ls())
```

# Activity Level Analysis
Now I collapse the data by activity. 
* NOTE: for research related to the actual time of day when people do particular 
activities, this is where you have to start fixing the code.
* NOTE: for location data, this is where you have to fix the code.
```{r labeling, echo = FALSE, include = FALSE, eval = FALSE}
load(file = "~/Desktop/TimeData/data/timeUseClean.rda")

# Summing activities by household and year to get average weekly hours while
# preserving all household information.
timeUse <- timeUse[, .(weeklyHours = sum(TUACTDUR24) / 60 * 7),
                   by = .(TUCASEID, TRTIER1P, TRTIER2P, TRCODEP, TEWHERE, GEMETSTA,
                          GTMETSTA, PEEDUCA, PEHSPNON, PTDTRACE, TEAGE, TELFS,
                          TESCHENR, TESCHLVL, TESEX, TESPEMPNOT, TRCHILDNUM,
                          TRDPFTPT, TRERNWA, TRYHHCHILD, TUDIARYDAY, TUFNWGTP,
                          TUYEAR, home, ageGroup)]

# Quality check for computed weekly hours
qualityCheck <- timeUse[, sum(weeklyHours), by = TUCASEID]

# Saving cleaned data
save(timeUse, file = "~/Desktop/TimeData/data/timeUseActivity.rda", compress = TRUE)
rm(list = ls())
```

Now I write a function that will flexibly allow analysis of trends in time use by group.
```{r aggFunction, eval = TRUE, warning = FALSE, echo = FALSE}
load(file = "~/Desktop/TimeData/data/timeUseActivity.rda")
activities <- levels(timeUse$TRTIER1P)
plots <- list()

codeColumn <- function(code) {
  if (nchar(code) %in% c(5, 6)) {
    codeLevel <- quote(TRCODEP)
  } else if (nchar(code) %in% c(3, 4)) {
    codeLevel <- quote(TRTIER2P)
  } else if (nchar(code) %in% c(1, 2)) {
    codeLevel <- quote(TRTIER1P)
  }
  return(codeLevel)
}

category <- function(groupName) {
    if (groupName == "sex") {
    lineCat <- c("TESEX")
  } else if (groupName == "age") {
    lineCat <- c("ageGroup")
  } else if (groupName == "ageSex") {
    lineCat <- c("TESEX", "ageGroup")
  } else if (groupName == "metro") {
    lineCat <- c("GTMETSTA")
  } else if (groupName == "sexMetro") {
    lineCat <- c("GTMETSTA", "TESEX")
  } else if (groupName == "ageMetro") {
    lineCat <- c("GTMETSTA", "ageGroup")
  } else {
    stop("Aggregation not implemented yet.")
  }
  group <- c("TUYEAR", lineCat)
  return(list(lineCat = lineCat, group = group))
}

plotGroup <- function(timeUse, code, groupName, title = NULL) {
  title <- ifelse(is.null(title), activities[as.numeric(substr(code, 1, 2))], title)
  
  # Matching input code with tier level of activity
  code <- ifelse(nchar(code) %% 2 == 0, as.character(code), paste0("0", code))
  substring(code, 1, 2) <- ifelse(substr(code, 1, 2) == "18", "17", "")
  substring(code, 1, 2) <- ifelse(substr(code, 1, 2) == "50", "18", "")
  codeLevel <- codeColumn(code)
  
  # Getting relevant aggregations
  temp <- category(groupName)
  lineCat <- temp$lineCat
  group <- temp$group
  
  # Aggregating to appropriate activity level
  tempData <- timeUse[as.numeric(eval(codeLevel)) == as.numeric(code),
                      .(weeklyHoursLevel = sum(weeklyHours)),
                      by = .(TUCASEID, TEWHERE, GEMETSTA, GTMETSTA, PEEDUCA, 
                             PEHSPNON, PTDTRACE, TEAGE, TELFS, TESCHENR, 
                             TESCHLVL, TESEX, TESPEMPNOT, TRCHILDNUM,
                             TRDPFTPT, TRERNWA, TRYHHCHILD, TUDIARYDAY, 
                             TUFNWGTP, TUYEAR, home, ageGroup)]

  # Weighted average hours per week spent on activity.
  graphData <- tempData[, .(meanTime = weighted.mean(weeklyHoursLevel, 
                                                     w = TUFNWGTP)),
                       by = eval(group)]

  plots <-  graphData %>% plot_ly(x = ~TUYEAR, y = ~meanTime, type = "scatter",
                                  mode = "lines", color = as.formula(paste0("~factor(", lineCat, ")"))) %>%
            layout(title = title,
                   xaxis = list(title = "Year"),
                   yaxis = list(title = "Weekly Hours"))
  return(plots)
}

# Function to generate subplot titles
fourTitles <- function(title1, title2, title3, title4) {
  annotations = list(
    list(
      x = 0.225, 
      y = 1.0, 
      font = list(size = 16), 
      showarrow = FALSE, 
      text = title1, 
      xanchor = "center", 
      xref = "paper", 
      yanchor = "bottom", 
      yref = "paper"
    ), 
    list(
      x = 0.775, 
      y = 1.0, 
      font = list(size = 16), 
      showarrow = FALSE, 
      text = title2, 
      xanchor = "center", 
      xref = "paper", 
      yanchor = "bottom", 
      yref = "paper"
    ), 
    list(
      x = 0.225, 
      y = 0.375, 
      font = list(size = 16), 
      showarrow = FALSE, 
      text = title3, 
      xanchor = "center", 
      xref = "paper", 
      yanchor = "bottom", 
      yref = "paper"
    ), 
    list(
      x = 0.775, 
      y = 0.375, 
      font = list(size = 16), 
      showarrow = FALSE, 
      text = title4, 
      xanchor = "center", 
      xref = "paper", 
      yanchor = "bottom", 
      yref = "paper"
    )
  )
  return(annotations)
}
```

## Trends in Time Use by Gender
Here are the graphs (left to right from top left: Personal Care; Household Activities; Caring for and Helping Household Members; Caring for and Helping non-Household members)
```{r genderPlotting, echo=TRUE, warning=FALSE}
plots <- lapply(c(1:16, 18, 50), function(x) plotGroup(timeUse, x, "sex"))
suppressWarnings(subplot(plots[1:4], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Personal Care",
                                                   "Household Activities",
                                                   "Caring for HH Members",
                                                   "Caring for non-HH Members")))
```

Here are the graphs (left to right from top left: Work; Education; Consumer Purchases; Professional and Personal Care Services)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[5:8], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Work",
                                                   "Education",
                                                   "Consumer Purchases",
                                                   "Professional Services")))
```

Here are the graphs (clockwise from top left: Household Services; Government Services; Eating and Drinking; Socializing, Relaxing and Leisure)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[9:12], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Household Services",
                                                   "Government Services",
                                                   "Eating and Drinking",
                                                   "Socializing, Relaxing, and Leisure")))
```

Here are the graphs (clockwise from top left: Sports, Exercise and Recreation; Religion; Volunteer; Telephone Calls)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[13:16], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Sports, Exercise, and Rec",
                                                   "Religion",
                                                   "Volunteer",
                                                   "Telephone Calls")))
```

Here is the last graph for travel and unknown
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[17:18], margin = 0.12, nrows = 1) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Travel", "Unknown", "", "")))
```

## Trends in Time Use by Age Group
The age groups are as follows: 18-24, 25-54, 55-64, 65+. 

Here are the graphs (left to right from top left: Personal Care; Household Activities; Caring for and Helping Household Members; Caring for and Helping non-Household members)
```{r agePlotting, echo=TRUE, warning=FALSE}
plots <- lapply(c(1:16, 18, 50), function(x) plotGroup(timeUse, x, "age"))
suppressWarnings(subplot(plots[1:4], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Personal Care",
                                                   "Household Activities",
                                                   "Caring for HH Members",
                                                   "Caring for non-HH Members")))
```

Here are the graphs (left to right from top left: Work; Education; Consumer Purchases; Professional and Personal Care Services)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[5:8], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Work",
                                                   "Education",
                                                   "Consumer Purchases",
                                                   "Professional Services")))
```

Here are the graphs (clockwise from top left: Household Services; Government Services; Eating and Drinking; Socializing, Relaxing and Leisure)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[9:12], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Household Services",
                                                   "Government Services",
                                                   "Eating and Drinking",
                                                   "Socializing, Relaxing, and Leisure")))
```

Here are the graphs (clockwise from top left: Sports, Exercise and Recreation; Religion; Volunteer; Telephone Calls)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[13:16], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Sports, Exercise, and Rec",
                                                   "Religion",
                                                   "Volunteer",
                                                   "Telephone Calls")))
```

Here is the last graph for travel and unknown
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[17:18], margin = 0.12, nrows = 1) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Travel", "Unknown", "", "")))
```

## Trends in Time Use by Age and Gender

Here are the graphs (left to right from top left: Personal Care; Household Activities; Caring for and Helping Household Members; Caring for and Helping non-Household members)
```{r ageSexPlotting, echo=TRUE, warning=FALSE}
plots <- lapply(c(1:16, 18, 50), function(x) plotGroup(timeUse, x, "metro"))
suppressWarnings(subplot(plots[1:4], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Personal Care",
                                                   "Household Activities",
                                                   "Caring for HH Members",
                                                   "Caring for non-HH Members")))
```

Here are the graphs (left to right from top left: Work; Education; Consumer Purchases; Professional and Personal Care Services)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[5:8], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Work",
                                                   "Education",
                                                   "Consumer Purchases",
                                                   "Professional Services")))
```

Here are the graphs (clockwise from top left: Household Services; Government Services; Eating and Drinking; Socializing, Relaxing and Leisure)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[9:12], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Household Services",
                                                   "Government Services",
                                                   "Eating and Drinking",
                                                   "Socializing, Relaxing, and Leisure")))
```

Here are the graphs (clockwise from top left: Sports, Exercise and Recreation; Religion; Volunteer; Telephone Calls)
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[13:16], margin = 0.12, nrows = 2) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Sports, Exercise, and Rec",
                                                   "Religion",
                                                   "Volunteer",
                                                   "Telephone Calls")))
```

Here is the last graph for travel and unknown
```{r, echo=TRUE, warning=FALSE}
suppressWarnings(subplot(plots[17:18], margin = 0.12, nrows = 1) %>% 
                   layout(title = "", 
                          annotations = fourTitles("Travel", "Unknown", "", "")))
```


# Location Analysis
First we collapse from the 6 digit activity level to the 2-digit activity level
by household and location to start conducting the analysis of where people 
are spending their time.
```{r, echo=FALSE, warning=FALSE}
load(file = "~/Desktop/TimeData/data/timeUseClean.rda")

# Summing time spent at each location for each two digit level activity by 
# household and year to get average weekly hours while preserving all 
# household information.
timeUse <- timeUse[, .(weeklyHours = sum(TUACTDUR24) / 60 * 7),
                   by = .(TUCASEID, TRTIER1P, TEWHERE, GEMETSTA,
                          GTMETSTA, PEEDUCA, PEHSPNON, PTDTRACE, TEAGE, TELFS,
                          TESCHENR, TESCHLVL, TESEX, TESPEMPNOT, TRCHILDNUM,
                          TRDPFTPT, TRERNWA, TRYHHCHILD, TUDIARYDAY, TUFNWGTP,
                          TUYEAR, home, ageGroup)]

# Saving cleaned data
save(timeUse, file = "~/Desktop/TimeData/data/timeUseLocation.rda", compress = TRUE)
rm(list = ls())
```

First, we'll just look at the share of time spent at each location by 2-digit activity
```{r}
load(file = "~/Desktop/TimeData/data/timeUseLocation.rda")
timeUse <- timeUse[, .(hours = weighted.mean(weeklyHours, w = TUFNWGTP)), 
                   by = .(TRTIER1P, TEWHERE, TUYEAR)]

timeUse[TRTIER1P == "Consumer Purchases"] %>% 
  group_by(TEWHERE) %>%
  plot_ly(x = ~TUYEAR, y = ~hours, color = ~TEWHERE)
```

